A Spectral Analysis of the Domain Decomposed Adjoint Neumann-Ulam Method

### 2012-11-26 doc/mandc2013/notes/spectral_theory.tex
updates to first mandc paper

  The domain decomposed behavior of the adjoint Neumann-Ulam method
  for solving linear operator equations is analyzed using the spectral
  properties of the linear operator. Relationships for the average
  length of the adjoint random walks and the fraction of histories
  leaking from a domain are made with respect to the Eigenvalues of
  the linear operator. In addition, a formal definition of the optical
  thickness of a domain in the decomposition is presented based on the
  spectral analysis. The one-speed, two-dimensional neutron diffusion
  equation is used as a model problem to test the spectral theory for
  symmetric operators. In general, the spectral theory shows good
  agreement with measured computational results.

@@

For problems where the linear operator is symmetric, a host of
analytic theories exist based on the eigenvalue spectrum of the
operator that characterize their behavior in the context of
deterministic linear solvers. Using past work, these theories are
adapted to the domain decomposed adjoint Neumann-Ulam method using the
one-speed, two-dimensional neutron diffusion equation. In this paper
we describe the adjoint Neumann-Ulam method followed by a presentation
and finite difference discretization of the model problem. Using the
linear system generated by this discretization, we use a spectral
theory to generate analytic relations for the eigenvalues of the
operator based on system parameters. Using the Eigenvalue spectra, we
then build relationships to characterize the transport of random walk
histories in a decomposed domain and compare these analytic results to
numerical experiments conducted with the model problem.

%%---------------------------------------------------------------------------%%
\section{The Adjoint Neumann-Ulam Method}
We seek solutions of the general linear problem in the following form:
\begin{equation}
  \ve{A} \ve{x} = \ve{b}\:,
  \label{eq:linear_problem}
\end{equation}
where $\ve{A} \in \mathbb{R}^{N \times N}$ is a linear operator such
that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
\mathbb{R}^N$ is the solution vector, and $\ve{b} \in \mathbb{R}^N$ is
the forcing term. Choosing the adjoint method Monte Carlo method to
invert the linear operator, we begin by defining the linear system
adjoint to Eq~(\ref{eq:linear_problem}):
\begin{equation}
  \ve{A}^T \ve{y} = \ve{d}\:,
  \label{eq:adjoint_linear_problem}
\end{equation}
where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
respectively and $\ve{A}^T$ is the adjoint operator.  With this
statement we can then define the split equation:
\begin{equation}
  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:,
  \label{eq:adjoint_split_system}
\end{equation}
where $\ve{H} = \ve{I} - \ve{A}$ and is defined as the
\textit{iteration matrix}.  Using these definitions, we can derive an
estimator from the adjoint method that will also give the solution
vector, $\ve{x}$. We can acquire the adjoint solution by forming the
Neumann series by writing Eq~(\ref{eq:adjoint_split_system}) as:
\begin{equation}
  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
  \label{eq:adjoint_split_system_2}
\end{equation}
which in turn yields the Neumann series using the adjoint operator:
\begin{equation}
  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
  \label{eq:adjoint_neumann_series}
\end{equation}
For this series to converge, we require that the spectral radius of
$\ve{H}$ must remain less than 1 as $\ve{H}^T$ contains the same
eigenvalues and therefore has the same spectral radius. We expand this
summation to again yield a series of transitions that can be
approximated by a Monte Carlo random walk sequence, this time forming
the Neumann series in reverse order:
\begin{equation}
  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
  \label{eq:adjoint_neumann_solution}
\end{equation}
We can readily build an estimator for the adjoint solution from this
series expansion, but we instead desire the direct solution. The
adjoint estimator can be related to the solution by defining the
following inner product equivalence \cite{spanier_monte_1969}:
\begin{equation}
  \langle \ve{A}^T \ve{x}, \ve{y} \rangle = \langle \ve{x}, \ve{A}
  \ve{y} \rangle\:.
  \label{eq:adjoint_operator_product}
\end{equation}
From this definition it follows that:
\begin{equation}
  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
  \label{eq:adjoint_vector_relation}
\end{equation}
Here we have 2 unknowns, $\ve{y}$ and $\ve{d}$ and therefore we
require two constraints to close the system. We use
Eq~(\ref{eq:adjoint_vector_relation}) as the first constraint and as
a second constraint we select:
\begin{equation}
  \ve{d} = \boldsymbol{\delta}_i\:,
  \label{eq:adjoint_second_constraint}
\end{equation}
where the $k^{th}$ component of the vector $\boldsymbol{\delta}_i$ is
the Dirac delta function $\delta_{i_k,i}$. If we apply
Eq~(\ref{eq:adjoint_second_constraint}) to our first constraint
Eq~(\ref{eq:adjoint_vector_relation}), we get the following convenient
outcome:
\begin{equation}
  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
  \boldsymbol{\delta}_i \rangle = x_i \:,
  \label{eq:inner_product_constraint}
\end{equation}
meaning that if we compute the inner product of the direct source and
the adjoint solution using a delta function source, we recover the
direct solution we are looking for

In terms of particle transport, this adjoint method is equivalent to a
traditional forward method. As a result of using the adjoint system,
we modify our probabilities and weights using the \textit{adjoint
  Neumann-Ulam decomposition} of $\ve{H}$:
\begin{equation}
  \ve{H}^{T} = \ve{P} \circ \ve{W}\:,
  \label{eq:adjoint_neumann_ulam}
\end{equation}
where now we are forming the decomposition with respect to the
transpose of $\ve{H}$. We then follow the same procedure as the direct
method for forming the probability and weight matrices in the
decomposition. Using the adjoint form, probabilities should instead be
column-scaled:
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,
  \label{eq:adjoint_probability}
\end{equation}
such that we expect to select a new state $j$ from the current state
in the random walk $j$ by sampling column-wise. Per
Eq~(\ref{eq:adjoint_neumann_ulam}), the transition weight is then
defined as:
\begin{equation}
  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
  \label{eq:adjoint_weight}
\end{equation}
Using the decomposition we can then define an expectation value for
the adjoint method. Using our result from
Eq~(\ref{eq:inner_product_constraint}) generated by applying the
adjoint constraints, the contribution to the solution in state $i$
from a particular random walk permutation is then:
\begin{equation}
  X_{\nu} = \sum_{m=0}^k W_{m} \delta_{i,i_m}\:,
  \label{eq:adjoint_permutation_contribution}
\end{equation}
where the Kronecker delta indicates that the tally contributes only in
the current state and $b_{i_0}$ will be the sampled source starting
weight. Note here that the estimator in
Eq~(\ref{eq:adjoint_permutation_contribution}) does not have a
dependency on the source state as in
Eq~(\ref{eq:adjoint_permutation_contribution}), providing a remedy for
the situation in the direct method where we must start a random walk
in each source state for every permutation such that we may compute a
contribution for that state. In the adjoint method, we instead tally
in all states and those of lesser importance will not be visited as
frequently by the random walk. Finally, the expectation value using
all permutations is:
\begin{equation}
  E\{X\} = \sum_{\nu} P_{\nu} X_{\nu}\:
  \label{eq:adjoint_expectation_value}
\end{equation}
which, if expanded in the same way as the direct method, directly
recovers the exact solution:
\begin{equation}
  \begin{split}
    E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N} b_{i_0} h_{i_0,i_1}h_{i_1,i_2}\ldots
    h_{i_{k-1},i_k} \delta_{i_k,j} \\ &= x_{j}\:,
  \end{split}
  \label{eq:adjoint_expectation_expansion}
\end{equation}
therefore also providing an unbiased Monte Carlo estimate of the
solution.

We also desire a criteria for random walk termination for problems
where only an approximate solution is necessary. For the adjoint
method, we utilize a \textit{relative weight cutoff}:
\begin{equation}
  W_f = W_c b_{i_0}\:,
  \label{eq:relative_weight_cutoff}
\end{equation}
where $W_c$ is defined as in the direct method. The adjoint random
walk will then be terminated after $m$ steps if $W_m < W_f$ as tally
contributions become increasingly small.

@@

We choose a finite difference scheme on a square Cartesian grid to
discretize the problem. For the Laplacian, we choose the 9-point
stencil shown in Figure~\ref{fig:stencil} over a grid of size $h$
\cite{leveque_finite_2007}:
\begin{multline}
  \nabla^2_9\phi = \frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
    + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20
    \phi_{i,j}]\:.
  \label{eq:nine_point_stencil}
\end{multline}
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.25}{\input{stencil.pdftex_t}}
  \end{center}
  \caption{\textbf{Nine-point Laplacian stencil.}}
  \label{fig:stencil}
\end{figure}
We then have the following linear system to solve:
\begin{multline}
  -\frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j} + 4
    \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ + \phi_{i-1,j+1}
    + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20 \phi_{i,j}] + \Sigma_a
  \phi_{i,j} = s_{i,j}\:,
  \label{eq:fd_system}
\end{multline}
and in operator form:
\begin{equation}
  \ve{D}\boldsymbol{\phi}=\ve{s}\:,
  \label{eq:operator_system}
\end{equation}
where $\ve{D}$ is the diffusion operator, $\ve{s}$ is the source in
vector form and $\boldsymbol{\phi}$ is the vector of unknown fluxes. We
note here that the diffusion operator is symmetric.

@@

As Eq~(\ref{eq:diffusion_eq}) is diagonally dominant, Jacobi
preconditioning is sufficient to reduce the spectral radius of the
iteration matrix below unity and therefore ensure convergence of the
Neumann series. The preconditioner in this case is then $\ve{M} =
diag(\ve{D})$ such that we are solving the following linear system:
\begin{equation}
  \ve{M}^{-1} \ve{D} \boldsymbol{\phi} = \ve{M}^{-1} \ve{s}\:.
  \label{eq:precond_diffsion}
\end{equation}
The operator $\ve{M}^{-1} \ve{D}$ is merely the original diffusion
operator with each row scaled by the diagonal component. As we have
defined a homogeneous domain, the scaling factor is the same for all
rows in the operator. This scaling factor, $\alpha$, is then defined
as the $\phi_{i,j}$ coefficient from Eq~(\ref{eq:fd_system}):
\begin{equation}
  \alpha = \Bigg[\frac{10 D}{3 h^2} + \Sigma_a\Bigg]^{-1}\:.
  \label{eq:jacobi_scaling}
\end{equation}
Using this coefficient, we then have the following spectrum of
preconditioned eigenvalues:
\begin{equation}
  \lambda_{p,q}(\ve{M}^{-1} \ve{D}) = \alpha \lambda_{p,q}(\ve{D})\:.
  \label{eq:preconditioned_eigenvalues}
\end{equation}

A good measure of how well-conditioned is a linear system can be
obtained by computing the condition number for the linear operator
defined as:
\begin{equation}
  \kappa(\ve{A}) = ||\ve{A}||\ ||\ve{A}^{-1}||\:,
  \label{eq:condition_number}
\end{equation}
which can be related to the Eigenvalues of the operator as:
\begin{equation}
  \kappa(\ve{A}) =
  \frac{\max_{p,q}\lambda_{p,q}(\ve{A})}{\min_{p,q}\lambda_{p,q}(\ve{A})}\:.
  \label{eq:condition_number_2}
\end{equation}
To compute the condition number for the preconditioned diffusion
system, we must therefore compute the minimum and maximum Eigenvalues
of the diffusion operator. Per Leveque's text
\cite{leveque_finite_2007}, we expect to find these minimum and
maximum eigenvalues when $p=q$. To find these values,
Eq~(\ref{eq:preconditioned_eigenvalues}) is plotted as a function of
$p$ with $p=q$ in Figure~\ref{fig:diffusion_spectrum} for various
values of $\Sigma_a$.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in,clip]{diffusion_spectrum.png}
  \end{center}
  \caption{\textbf{Eigenvalue spectra for the diffusion equation.}}
  \label{fig:diffusion_spectrum}
\end{figure}
We see a minimum eigenvalue at $p=0$ and a maximum at $p=10$, letting
us compute the condition number for the Jacobi preconditioned
diffusion operator:
\begin{equation}
  \kappa(\ve{M}^{-1} \ve{D}) =
  \frac{\frac{D}{6h^2}\Sigma_a}{\frac{D}{3h^2}[8 \cos(10 \pi h) + 2
      \cos^2(10 \pi h) - 10] + \Sigma_a}\:.
  \label{eq:diffusion_condition}
\end{equation}
Visually we can also see how increasing the absorption in the problem
creates a more well conditioned system as less diffusion is
expected. If we have a problem with only scattering, $\Sigma_a = 0$
and we observe an infinite condition number as $\lambda_{min}
\rightarrow 0$. We note also that the diffusion operator contains
purely positive Eigenvalues and can therefore be classified as
symmetric positive definite (SPD).

\subsection{Iteration Matrix Spectrum}
The spectral radius of the iteration matrix is obtained by seeking its
largest Eigenvalue. As with the diffusion operator, we can use the
same analysis techniques to find the Eigenvalues for the iteration
matrix. We use a few simplifications by noting that if the Jacobi
preconditioned spectral radius is $\ve{H} = \ve{I} -
\ve{M}^{-1}\ve{D}$, the we expect all terms on the diagonal of the
iteration matrix to be zero such that we have the following stencil:
\begin{equation}
  \ve{H}\boldsymbol{\phi} = -\frac{\alpha D}{6h^2}[4 \phi_{i-1,j}
    + 4 \phi_{i+1,j} + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} +
    \phi_{i-1,j-1}\\ + \phi_{i-1,j+1} + \phi_{i+1,j-1} +
    \phi_{i+1,j+1}]\:.
  \label{eq:iteration_stencil}
\end{equation}
Inserting the Eigenfunctions defined by Eq~\ref{eq:eigenfunction_form}
we get:
\begin{multline}
  \lambda_{p,q}(\ve{H}) = -\frac{\alpha D}{6h^2}\Big[4 e^{-2 \pi \imath p
      h} + 4 e^{2 \pi \imath p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2
      \pi \imath q h} + e^{-2 \pi \imath p h} e^{-2 \pi \imath q h}
    \\ + e^{-2 \pi \imath p h} e^{2 \pi \imath q q} + e^{2 \pi \imath
      p h} e^{-2 \pi \imath q h} + e^{2 \pi \imath p h} e^{2 \pi
      \imath q h}\Big]\:,
  \label{eq:iteration_deriv}
\end{multline}
which simplifies to:
\begin{equation}
  \lambda_{p,q}(\ve{H}) = -\frac{\alpha D}{6h^2}[ 8 \cos(\pi p h) + 8
    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h)]\:,
  \label{eq:iteration_spectrum}
\end{equation}
giving the Eigenvalue spectrum for the Jacobi preconditioned iteration
matrix. As with the linear operator, we expect the maximum Eigenvalue
to exist when $p=q$. In addition, we note that we have the same cosine
form of the spectrum as for the linear operator expect that the
spectra will be shifted by 10 in the $-p$ direction. We therefore
expect the maximum Eigenvalues to be found when $p=q=0$, giving the
following for the spectral radius of the Jacobi precondition iteration
matrix:
\begin{equation}
  \rho(\ve{H}) = \frac{10 \alpha D}{3 h^2}\:.
  \label{eq:iteration_radius}
\end{equation}

>>>>

  The domain decomposed behavior of the adjoint Neumann-Ulam method
  for solving linear operator equations is analyzed using the spectral
  properties of the linear operator. Relationships for the average
  length of the adjoint random walks and the fraction of histories
  leaking from a domain in the decomposition are made with respect to
  the Eigenvalues and other properties of the linear operator. In
  addition, a formal definition of the optical thickness of a domain
  in the decomposition is presented based on the spectral
  analysis. The one-speed, two-dimensional neutron diffusion equation
  is used as a model problem to test the spectral theory for symmetric
  operators. In general, the spectral theory shows good agreement with
  measured computational results.

@@

For problems where the linear operator is symmetric, a host of
analytic theories exist based on the Eigenvalue spectrum of the
operator that characterize their behavior in the context of
deterministic linear solvers. Using past work, these theories are
adapted to the domain decomposed adjoint Neumann-Ulam method using the
one-speed, two-dimensional neutron diffusion equation. In this paper
we describe the adjoint Neumann-Ulam method followed by a presentation
and finite difference discretization of the model problem. Using the
linear system generated by this discretization, we use a spectral
theory to generate analytic relations for the Eigenvalues of the
operator based on system parameters. Using the Eigenvalue spectra, we
then build relationships to characterize the transport of random walk
histories in a decomposed domain and compare these analytic results to
numerical experiments conducted with the model problem.

%%---------------------------------------------------------------------------%%
\section{The Adjoint Neumann-Ulam Method}
We seek solutions of the general linear problem in the following form:
\begin{equation}
  \ve{A} \ve{x} = \ve{b}\:,
  \label{eq:linear_problem}
\end{equation}
where $\ve{A} \in \mathbb{R}^{N \times N}$ is a linear operator such
that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
\mathbb{R}^N$ is the solution vector, and $\ve{b} \in \mathbb{R}^N$ is
the forcing term. Choosing the adjoint method Monte Carlo method to
invert the linear operator, we begin by defining the linear system
adjoint to Eq~(\ref{eq:linear_problem}):
\begin{equation}
  \ve{A}^T \ve{y} = \ve{d}\:,
  \label{eq:adjoint_linear_problem}
\end{equation}
where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
respectively and $\ve{A}^T$ is the adjoint operator.  With this
statement we can then define the split equation:
\begin{equation}
  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:,
  \label{eq:adjoint_split_system}
\end{equation}
where $\ve{H} = \ve{I} - \ve{A}$ and is defined as the
\textit{iteration matrix}.  Using these definitions, we can derive an
estimator from the adjoint method that will also give the solution
vector, $\ve{x}$. We can acquire the adjoint solution by forming the
Neumann series by writing Eq~(\ref{eq:adjoint_split_system}) as:
\begin{equation}
  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
  \label{eq:adjoint_split_system_2}
\end{equation}
which in turn yields the Neumann series using the adjoint operator:
\begin{equation}
  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
  \label{eq:adjoint_neumann_series}
\end{equation}
For this series to converge, we require that the spectral radius of
$\ve{H}$ must remain less than 1 as $\ve{H}^T$ contains the same
Eigenvalues and therefore has the same spectral radius. We expand this
summation to again yield a series of transitions that can be
approximated by a Monte Carlo random walk sequence, forming the
Neumann series in reverse order:
\begin{equation}
  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
  \label{eq:adjoint_neumann_solution}
\end{equation}
We can readily build an estimator for the adjoint solution from this
series expansion, but we instead desire the solution to
Eq~(\ref{eq:linear_problem}). The adjoint estimator can be related to
the solution by defining the following inner product equivalence
\cite{spanier_monte_1969}:
\begin{equation}
  \langle \ve{A}^T \ve{x}, \ve{y} \rangle = \langle \ve{x}, \ve{A}
  \ve{y} \rangle\:.
  \label{eq:adjoint_operator_product}
\end{equation}
From this definition it follows that:
\begin{equation}
  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
  \label{eq:adjoint_vector_relation}
\end{equation}
Here we have 2 unknowns, $\ve{y}$ and $\ve{d}$, and therefore we
require two constraints to close the system. We use
Eq~(\ref{eq:adjoint_vector_relation}) as the first constraint and as a
second constraint we select:
\begin{equation}
  \ve{d} = \boldsymbol{\delta}_i\:,
  \label{eq:adjoint_second_constraint}
\end{equation}
where the $k^{th}$ component of the vector $\boldsymbol{\delta}_i$ is
the Dirac delta function $\delta_{i_k,i}$. If we apply
Eq~(\ref{eq:adjoint_second_constraint}) to our first constraint
Eq~(\ref{eq:adjoint_vector_relation}), we get the following convenient
outcome:
\begin{equation}
  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
  \boldsymbol{\delta}_i \rangle = x_i \:,
  \label{eq:inner_product_constraint}
\end{equation}
meaning that if we compute the inner product of the original source and
the adjoint solution using a delta function source, we recover the
original solution we are looking for. In terms of particle transport, a
delta function source means that this adjoint method is equivalent to
a traditional forward method where tallies are made only in the state
in which the random walk currently resides.

In order to perform the random walks required to approximate the
summation in Eq~(\ref{eq:adjoint_neumann_solution}), probabilities and
weights are generated using the \textit{adjoint Neumann-Ulam
  decomposition} of $\ve{H}$:
\begin{equation}
  \ve{H}^{T} = \ve{P} \circ \ve{W}\:,
  \label{eq:adjoint_neumann_ulam}
\end{equation}
where now we are forming the decomposition with respect to the
transpose of $\ve{H}$. We then build the probability and weight
matrices in the decomposition where probabilities are column-scaled:
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,
  \label{eq:adjoint_probability}
\end{equation}
such that we expect to select a new state $j$ from the current state
in the random walk $j$ by sampling column-wise. Per
Eq~(\ref{eq:adjoint_neumann_ulam}), the transition weight is then
defined as:
\begin{equation}
  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
  \label{eq:adjoint_weight}
\end{equation}
Using the decomposition we can then define an expectation value for
the adjoint method. Using our result from
Eq~(\ref{eq:inner_product_constraint}) generated by applying the
adjoint constraints, the contribution to the solution in state $i$
from a particular random walk permutation is then:
\begin{equation}
  X_{\nu} = \sum_{m=0}^k W_{m} \delta_{i,i_m}\:,
  \label{eq:adjoint_permutation_contribution}
\end{equation}
where the Kronecker delta indicates that the tally contributes only in
the current state and $b_{i_0}$ will be the sampled source starting
weight. Finally, the expectation value using all permutations is:
\begin{equation}
  E\{X\} = \sum_{\nu} P_{\nu} X_{\nu}\:
  \label{eq:adjoint_expectation_value}
\end{equation}
which if expanded directly recovers the exact solution:
\begin{equation}
  \begin{split}
    E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N} b_{i_0} h_{i_0,i_1}h_{i_1,i_2}\ldots
    h_{i_{k-1},i_k} \delta_{i_k,j} \\ &= x_{j}\:,
  \end{split}
  \label{eq:adjoint_expectation_expansion}
\end{equation}
therefore also providing an unbiased Monte Carlo estimate of the
solution.

We also desire a criteria for random walk termination for problems
where only an approximate solution is necessary. For the adjoint
method, we utilize a relative weight cutoff parameter:
\begin{equation}
  W_f = W_c b_{i_0}\:,
  \label{eq:relative_weight_cutoff}
\end{equation}
where $W_c$ is the \textit{weight cutoff}. The adjoint random
walk will then be terminated after $m$ steps if $W_m < W_f$ as tally
contributions become increasingly small.

@@

We choose a finite difference scheme on a square Cartesian grid to
discretize the problem. For the Laplacian, we choose the 9-point
stencil shown in Figure~\ref{fig:stencil} over a grid of size $h$
\cite{leveque_finite_2007}:
\begin{multline}
  \nabla^2_9\phi = \frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
    + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20
    \phi_{i,j}]\:.
  \label{eq:nine_point_stencil}
\end{multline}
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.25}{\input{stencil.pdftex_t}}
  \end{center}
  \caption{\textbf{Nine-point Laplacian stencil.}}
  \label{fig:stencil}
\end{figure}
We then have the following linear system to solve:
\begin{multline}
  -\frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j} + 4
    \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ + \phi_{i-1,j+1}
    + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20 \phi_{i,j}] + \Sigma_a
  \phi_{i,j} = s_{i,j}\:,
  \label{eq:fd_system}
\end{multline}
and in operator form:
\begin{equation}
  \ve{D}\boldsymbol{\phi}=\ve{s}\:,
  \label{eq:operator_system}
\end{equation}
where $\ve{D}$ is the diffusion operator, $\ve{s}$ is the source in
vector form and $\boldsymbol{\phi}$ is the vector of unknown fluxes.

@@

As Eq~(\ref{eq:diffusion_eq}) is diagonally dominant, Jacobi
preconditioning is sufficient to reduce the spectral radius of the
iteration matrix below unity and therefore ensure convergence of the
Neumann series. The preconditioner in this case is then $\ve{M} =
diag(\ve{D})$ such that we are solving the following linear system:
\begin{equation}
  \ve{M}^{-1} \ve{D} \boldsymbol{\phi} = \ve{M}^{-1} \ve{s}\:.
  \label{eq:precond_diffsion}
\end{equation}
The operator $\ve{M}^{-1} \ve{D}$ is merely the original diffusion
operator with each row scaled by the diagonal component. As we have
defined a homogeneous domain, the scaling factor is the same for all
rows in the operator. This scaling factor, $\alpha$, is then defined
as the $\phi_{i,j}$ coefficient from Eq~(\ref{eq:fd_system}):
\begin{equation}
  \alpha = \Bigg[\frac{10 D}{3 h^2} + \Sigma_a\Bigg]^{-1}\:.
  \label{eq:jacobi_scaling}
\end{equation}
Using this coefficient, we then have the following spectrum of
preconditioned Eigenvalues:
\begin{equation}
  \lambda_{p,q}(\ve{M}^{-1} \ve{D}) = \alpha \lambda_{p,q}(\ve{D})\:.
  \label{eq:preconditioned_Eigenvalues}
\end{equation}

A good measure of how well-conditioned is a linear system can be
obtained by computing the condition number for the linear operator
defined as:
\begin{equation}
  \kappa(\ve{A}) = ||\ve{A}||\ ||\ve{A}^{-1}||\:,
  \label{eq:condition_number}
\end{equation}
which can be related to the Eigenvalues of the operator as:
\begin{equation}
  \kappa(\ve{A}) =
  \frac{\max_{p,q}\lambda_{p,q}(\ve{A})}{\min_{p,q}\lambda_{p,q}(\ve{A})}\:.
  \label{eq:condition_number_2}
\end{equation}
To compute the condition number for the preconditioned diffusion
system, we must therefore compute the minimum and maximum Eigenvalues
of the diffusion operator. Per Leveque's text
\cite{leveque_finite_2007}, we expect to find these minimum and
maximum Eigenvalues when $p=q$. To find these values,
Eq~(\ref{eq:preconditioned_Eigenvalues}) is plotted as a function of
$p$ with $p=q$ in Figure~\ref{fig:diffusion_spectrum} for various
values of $\Sigma_a$.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in,clip]{diffusion_spectrum.png}
  \end{center}
  \caption{\textbf{Eigenvalue spectra for the diffusion equation.}}
  \label{fig:diffusion_spectrum}
\end{figure}
We see a minimum Eigenvalue at $p=0$ and a maximum at $p=10$, letting
us compute the condition number for the Jacobi preconditioned
diffusion operator:
\begin{equation}
  \kappa(\ve{M}^{-1} \ve{D}) =
  \frac{\frac{D}{6h^2}\Sigma_a}{\frac{D}{3h^2}[8 \cos(10 \pi h) + 2
      \cos^2(10 \pi h) - 10] + \Sigma_a}\:.
  \label{eq:diffusion_condition}
\end{equation}
Visually we can also see how increasing the absorption in the problem
creates a more well conditioned system as less diffusion is
expected. If we have a problem with only scattering, $\Sigma_a = 0$
and we observe an infinite condition number as $\lambda_{min}
\rightarrow 0$.

\subsection{Iteration Matrix Spectrum}
The spectral radius of the iteration matrix is obtained by seeking its
largest Eigenvalue. As with the diffusion operator, we can use the
same analysis techniques to find the Eigenvalues for the iteration
matrix. We use a few simplifications by noting that if the Jacobi
preconditioned iteration matrix is $\ve{H} = \ve{I} -
\ve{M}^{-1}\ve{D}$, the we except all terms on the diagonal of the
iteration matrix to be zero such that we have the following stencil:
\begin{equation}
  \ve{H}\boldsymbol{\phi} = -\frac{\alpha D}{6h^2}[4 \phi_{i-1,j}
    + 4 \phi_{i+1,j} + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} +
    \phi_{i-1,j-1}\\ + \phi_{i-1,j+1} + \phi_{i+1,j-1} +
    \phi_{i+1,j+1}]\:.
  \label{eq:iteration_stencil}
\end{equation}
Inserting the Eigenfunctions defined by Eq~\ref{eq:eigenfunction_form}
we get:
\begin{multline}
  \lambda_{p,q}(\ve{H}) = -\frac{\alpha D}{6h^2}\Big[4 e^{-2 \pi \imath p
      h} + 4 e^{2 \pi \imath p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2
      \pi \imath q h} + e^{-2 \pi \imath p h} e^{-2 \pi \imath q h}
    \\ + e^{-2 \pi \imath p h} e^{2 \pi \imath q q} + e^{2 \pi \imath
      p h} e^{-2 \pi \imath q h} + e^{2 \pi \imath p h} e^{2 \pi
      \imath q h}\Big]\:,
  \label{eq:iteration_deriv}
\end{multline}
which simplifies to:
\begin{equation}
  \lambda_{p,q}(\ve{H}) = -\frac{\alpha D}{6h^2}[ 8 \cos(\pi p h) + 8
    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h)]\:,
  \label{eq:iteration_spectrum}
\end{equation}
giving the Eigenvalue spectrum for the Jacobi preconditioned iteration
matrix. As with the linear operator, we expect the maximum Eigenvalue
to exist when $p=q$. In addition, we note that we have the same cosine
form of the spectrum as for the linear operator expect that the
spectra will be shifted by 10 in the $-p$ direction. We therefore
expect the maximum Eigenvalues to be found when $p=q=0$, giving the
following for the spectral radius of the Jacobi precondition iteration
matrix:
\begin{equation}
  \rho(\ve{H}) = \frac{10 \alpha D}{3 h^2}\:.
  \label{eq:iteration_radius}
\end{equation}
